---
title: "Untitled"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
# import libraries
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(glmnet)
library(ggplot2)
library(GEOquery)
library(Biobase)
```

# Prepare data
## Query and parse data from GEO
```{r}
gse40279 <- getGEO(GEO = "GSE40279", destdir = "data", GSEMatrix = TRUE, parseCharacteristics = TRUE)
```

## Extract data
```{r}
eset40279 <- gse40279[[1]]
pheno40279 <- pData(eset40279)
feature40279 <- fData(eset40279)
expr40279 <- exprs(eset40279)
```

## Save for future use 
```{r}
write.csv(pheno40279, file = "data/pheno40279")
# write.csv(feature40279, file = "data/feature40279")
write.csv(expr40279, file = "data/expr40279")
```

## Extract relavent features
```{r}
coldata <- pheno40279 %>%
  select(c("geo_accession", "age (y):ch1", "ethnicity:ch1", "gender:ch1")) %>%
  rename(
    sample_id = geo_accession,
    age = `age (y):ch1`,
    ethnicity = `ethnicity:ch1`,
    gender = `gender:ch1`
  ) %>%
  mutate(age = as.numeric(age))
head(coldata, n = 10)
```

## View data
```{r}
expr40279 <- t(expr40279)
expr40279 <- as.data.frame(expr40279)
head(expr40279, n = 10)
```

## merge selected feature with methylation data
```{r}
expr40279$sample_id <- rownames(expr40279)
expr40279 <- expr40279 %>%
  relocate(sample_id, .before = 1)
merged40279 <- inner_join(coldata, expr40279, by = "sample_id")
head(merged40279, n = 10)
```

## visualize data
### distribution by age
```{r}
ggplot(data = merged40279, mapping = aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "black", color = "white") +
  labs(title = "Distribution of Age",
       x = "Age (years)",
       y = "Count (# of sample)") +
  scale_x_continuous(breaks = seq(20, 100, by = 20)) +
  scale_y_continuous(breaks = seq(0, 100, by = 20)) +
  theme_minimal()
```

### perform PCA
```{r}
pca40279_res <- prcomp(merged40279[, -(1:4)], center = TRUE, scale. = TRUE)
summary(pca40279_res)
pca40279_pc1_pc2 <- as.data.frame(pca40279_res$x[, 1:2])
pca40279_pc1_pc2 <- bind_cols(pca40279_pc1_pc2, merged40279[, c("sample_id","age", "ethnicity", "gender")])
```

### plot PCA
```{r}
ggplot(pca40279_pc1_pc2, aes(x = PC1, y = PC2, color = age)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_gradient(low = "blue", high = "red") +
  labs(title = "PCA of Methylation Data by Age") +
  theme_minimal()
```



## split data into training 75% and testing sets 25%
```{r}
set.seed(123)
# First, define age bins (quartiles or custom)
merged40279 <- merged40279 %>%
  mutate(age_bin = ntile(age, 4))  # split age into 4 quantile bins

# Check how many samples per group
table(merged40279$age_bin, merged40279$gender)

train_data <- merged40279 %>%
  group_by(age_bin, gender) %>%
  sample_frac(0.75)

# Remaining 25% go to test set
test_data <- anti_join(merged40279, train_data, by = "sample_id")
```

```{r}
# Check age distribution
summary(train_data$age)
summary(test_data$age)

# Check gender balance
table(train_data$gender)
table(test_data$gender)
```


# Fit model
## apply Elastic Net algorithm in R package glmnet (have those CpGs contribute less to prediction excluded by have their coefficients to or close to zero, thus reduce the number of variable)
```{r}

```

## 10-fold crossvalidation to optimize regularization parameters
```{r}

```

## bootstrap 500 times, build model for each bootstrap
```{r}

```

## markers presented in more than half of all bootstraps were included in the final model
```{r}

```














































